---
title: "CUNY Summer Bridge R Week 3 Lab"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.Load the diamonds dataset, which is included with the ggplot2 package. Identify which variables in the
diamond set are numeric, and which are categorical (factors).
```{r}
install.packages("ggplot2")
library("ggplot2")
data("diamonds")
view(diamonds)
class(diamonds$carat) #Numeric
class(diamonds$cut) #Ordered Factor
class(diamonds$color) #Ordered Factor
class(diamonds$clarity) #Ordered Factor
class(diamonds$depth) #Numeric
class(diamonds$table) #Numeric
class(diamonds$price) #Integer
class(diamonds$x) #Numeric
class(diamonds$y) #Numeric
class(diamonds$z) #Numeric
```

2. Generate summary level descriptive statistics: Show the mean, median, 25th and 75th quartiles, min, and
max for each of the applicable variables in diamonds.
```{r}
summary(diamonds)
```

3. Determine the frequency for each of the diamond colors.
```{r}
D <- 0
E <- 0
F <- 0
G <- 0
H <- 0
I <- 0
J <- 0
for (i in 1:length(diamonds$color)){
  if (diamonds$color[i]=="D"){
    D <- D+1
  } else if (diamonds$color[i]=="E"){
    E <- E+1
  } else if (diamonds$color[i]=="F"){
    F <- F+1
  } else if (diamonds$color[i]=="G"){
    G <- G+1
  } else if (diamonds$color[i]=="H"){
    H <- H+1
  } else if (diamonds$color[i]=="I"){
    I <- I+1
  } else if (diamonds$color[i]=="J"){
    J <- J+1
  } 
}
```

4. Determine the frequency for each of the diamond cuts, by color.
```{r}
table(diamonds$color, diamonds$cut)
```

5. Create a graph for a single numeric variable.
```{r}
boxplot(diamonds$price)
qplot(price, data=diamonds)
```

6. Create a scatterplot of two numeric variables.
```{r}
plot(diamonds$price ~ diamonds$carat)
qplot(carat, price, data=diamonds)
```